---
title: 'Create Random Samples'
output:
  html_document:
    code_folding: 'hide'
---

```{r load_packages, echo=TRUE, include=FALSE}
library(tidyverse) ; library(reshape2) ; library(glue) ; library(plotly)
library(RColorBrewer) ; library(viridis) ; require(gridExtra)
library(DESeq2) ; library(sva)
library(foreach)
library(doParallel)
library(expss)
library(knitr)
```

Since we are going to create many random samples, it would be a lot of work to manually filter each one of them, so instead I'm going to use the already filtered dataset obtained from preprocessing all the brain regions together and just select different columns for each sample

```{r prepare_data}
# Load Filtered raw data from the experiment including all brain regions
load('./../../AllRegions/Data/filtered_raw_data.RData')

# Create original DESeqDataSet object
counts = datExpr %>% as.matrix
rowRanges = GRanges(datGenes$chromosome_name,
                  IRanges(datGenes$start_position, width=datGenes$length),
                  strand=datGenes$strand,
                  feature_id=datGenes$ensembl_gene_id)
se = SummarizedExperiment(assays=SimpleList(counts=counts), rowRanges=rowRanges, colData=datMeta)
dds = DESeqDataSet(se, design = ~ Diagnosis)
dds = estimateSizeFactors(dds)
norm.cts = counts(dds, normalized=TRUE)

# Estimate surrogate variables
mod = model.matrix(~ Diagnosis, colData(dds))
mod0 = model.matrix(~ 1, colData(dds))
sva_fit = svaseq(norm.cts, mod=mod, mod0=mod0)
sv_data = sva_fit$sv %>% data.frame
colnames(sv_data) = paste0('SV',1:ncol(sv_data))
datMeta = cbind(datMeta, sv_data)


rm(mod, mod0, norm.cts, sv_data, sva_fit, se, counts, rowRanges)
```


The dataset consisting of all the brain regions except for the Occipital Lobe contains 59 samples, so, to make my samples comparable, I'm going to create random samples of size 59

```{r}

n = 10000
set.seed(123)
samples_mat = replicate(n, sort(sample(1:nrow(datMeta), size = 59)))

```

```{r perform_samplings}

if(!file.exists('./../Data/random_samples_DE.RData')){
  get_DE_info_from_sample = function(s){
    
    # Craete DatExpr and datMeta for this sample
    datExpr_sample = datExpr[,s]
    datMeta_sample = datMeta[s,]
    
    # Create DESeqDataSet object for the sample
    counts = datExpr_sample %>% as.matrix
    rowRanges = GRanges(datGenes$chromosome_name, IRanges(datGenes$start_position, width=datGenes$length),
                        strand=datGenes$strand, feature_id=datGenes$ensembl_gene_id)
    se = SummarizedExperiment(assays=SimpleList(counts=counts), rowRanges=rowRanges, colData=datMeta_sample)
    dds = DESeqDataSet(se, design = ~ Batch + SV1 + SV2 + SV3 + SV4 + SV5 + SV6 + SV7 + SV8 + SV9 + 
                                      SV10 + SV11 + SV12 + SV13 + Diagnosis)
    # Perform DEA
    dds = DESeq(dds, quiet=TRUE)
    DE_info = results(dds, lfcThreshold=0, altHypothesis='greaterAbs') %>% data.frame
    
    output = data.frame('ID'=rownames(DE_info), 'lfc'=DE_info$log2FoldChange, 'padj'=DE_info$padj)
    
    return(output)
  }
  
  
  #setup parallel backend to use many processors
  cores = detectCores()
  cl = makeCluster(cores[1]-1) #not to overload your computer
  registerDoParallel(cl)

  output_matrix = foreach(i=1:n, .combine=cbind) %dopar% {
    
    library(dplyr) ; library(DESeq2)
    
    s = samples_mat[,i] %>% as.vector
    DE_info_sample = get_DE_info_from_sample(s)
    colnames(DE_info_sample) = c('ID', paste0('lfc_s',i), paste0('padj_s',i))
    
    temp_matrix = data.frame('ID' = rownames(datExpr)) %>% left_join(DE_info_sample, by = 'ID')
    if(i>1) temp_matrix = temp_matrix %>% dplyr::select(-ID)
    
    return(temp_matrix)
  }
  
  lfc_mat = output_matrix %>% dplyr::select(ID, dplyr::contains('lfc'))
  padj_mat = output_matrix %>% dplyr::select(ID, dplyr::contains('padj'))
  
  signif_mat = cbind(as.character(padj_mat$ID), padj_mat[,-1]<0.05) %>% data.frame
  signif_mat[is.na(signif_mat)] = FALSE
  signif_mat[,-1] = sapply(signif_mat[,-1], as.logical)
  colnames(signif_mat) = c('ID', gsub('padj','signif',colnames(signif_mat)[-1]))
  
  # Save data
  save(samples_mat, lfc_mat, padj_mat, signif_mat, file='./../Data/random_samples_DE.RData')
  
  rm(get_DE_info_from_sample)
} else {
  load('./../Data/random_samples_DE.RData')
}
```

#### Total number of DE genes on each random sample
```{r}

plot_data = data.frame('random_sample' = 1:(ncol(signif_mat)-1), 'tot_signif_genes' = colSums(signif_mat[,-1]))

summary(plot_data$tot_signif_genes)

ggplotly(plot_data %>% ggplot(aes(tot_signif_genes)) + 
         geom_histogram(alpha=0.5, color='#009999', fill='#009999', binwidth = 300) + 
         geom_vline(xintercept=mean(plot_data$tot_signif_genes), color = 'gray') + 
         xlab('Total number of DE genes in random sample') + theme_minimal())

```

#### Number of times each gene was found to be DE
```{r, warning=FALSE, message=FALSE}
plot_data = data.frame('Gene' = signif_mat$ID, 'n_times_DE' = rowSums(signif_mat[,-1])) %>%
            arrange(n_times_DE)

summary(plot_data$n_times_DE)

ggplotly(plot_data %>% ggplot(aes(n_times_DE)) + geom_histogram(alpha=0.5, color='#cc0066', fill='#cc0066') + 
         geom_vline(xintercept=mean(plot_data$n_times_DE), color = 'gray') + 
         xlab('Number of times each gene was found to be DE in a sample') + theme_minimal())

```

#### Number of genes found to be DE in at least n random samples

```{r}
plot_data = plot_data %>% group_by(n_times_DE) %>% tally() %>% arrange(desc(n_times_DE)) %>% 
            mutate('cum_DE_genes' = cumsum(n)) %>% arrange(n_times_DE)

ggplotly(plot_data %>% ggplot(aes(n_times_DE, cum_DE_genes)) + geom_point(shape='o', alpha=1, color='#cc0066') + 
         xlab('No. times a gene is found DE') + ylab('Number of genes found to be DE at least that number of times') + 
         theme_minimal())
```



