---
title: 'Exploratory Analysis for Frontal Lobe'
output:
  html_document:
    code_folding: 'hide'
---

```{r load_packages, echo=TRUE, warning=FALSE, message=FALSE}
#setwd('/afs/inf.ed.ac.uk/user/s17/s1725186/Documents/PhD-Models/FirstPUModel/RMarkdowns')

library(tidyverse) ; library(reshape2) ; library(glue) ; library(plotly) ; library(plotlyutils)
library(RColorBrewer) ; library(viridis) ; require(gridExtra) ; library(GGally)
library(Rtsne)
library(ClusterR)
library(DESeq2) ; library(biomaRt)
library(knitr)
```

Load preprocessed dataset (preprocessing code in 20_02_21_data_preprocessing.Rmd)
```{r load_dataset, echo=TRUE, message=FALSE, warning=FALSE}
# Gandal dataset
load('./../Data/preprocessed_data.RData')
datExpr = datExpr %>% data.frame
DE_info = DE_info %>% data.frame

# GO Neuronal annotations: regex 'neuron' in GO functional annotations and label the genes that make a match as neuronal
GO_annotations = read.csv('./../Data/genes_GO_annotations.csv')
GO_neuronal = GO_annotations %>% filter(grepl('neuron', go_term)) %>% 
              mutate('ID'=as.character(ensembl_gene_id)) %>% 
              dplyr::select(-ensembl_gene_id) %>% distinct(ID) %>%
              mutate('Neuronal'=1)

# Update DE_info with Neuronal information
DE_info = DE_info %>% mutate('ID'=rownames(.)) %>% left_join(GO_neuronal, by='ID') %>%
  mutate(Neuronal=ifelse(is.na(Neuronal), 0, Neuronal)) %>%
  mutate(significant=padj<0.05 & !is.na(padj))

rm(GO_annotations)
```
<br>

### All samples together

- There seem to be two different behaviours in mean expression by gene

```{r, fig.width=10}
plot_data = data.frame('ID'=rownames(datExpr), 'Mean'=rowMeans(datExpr))
p1 = ggplotly(plot_data %>% ggplot(aes(Mean)) + geom_density(color='#0099cc', fill='#0099cc', alpha=0.3) +
              scale_x_log10() + theme_minimal())

plot_data = data.frame('ID'=colnames(datExpr), 'Mean'=colMeans(datExpr))
p2 = ggplotly(plot_data %>% ggplot(aes(Mean)) + geom_density(color='#0099cc', fill='#0099cc', alpha=0.3) +
              theme_minimal() + ggtitle('Mean expression density by Gene (left) and by Sample (right)'))

subplot(p1, p2, nrows=1)

rm(p1, p2, plot_data)
```

#### Grouping genes by Neuronal tag and samples by Diagnosis

- The two groups of genes seem to be partially characterised by genes with Neuronal function

- In general, the autism group has a bigger mean and is more spread out than the control group

```{r GO_annotations, warning=FALSE, fig.width=10}
plot_data = data.frame('ID'=rownames(datExpr), 'Mean'=rowMeans(datExpr)) %>% 
            left_join(GO_neuronal, by='ID') %>% mutate('Neuronal'=ifelse(is.na(Neuronal),F,T))
p1 = plot_data %>% ggplot(aes(Mean, color=Neuronal, fill=Neuronal)) + geom_density(alpha=0.3) +
                   scale_x_log10() + theme_minimal() + theme(legend.position='bottom') + 
                   ggtitle('Mean expression density by gene')

plot_data = data.frame('ID'=colnames(datExpr), 'Mean'=colMeans(datExpr)) %>% 
            mutate('ID'=substring(ID,2)) %>% left_join(datMeta, by=c('ID'='Dissected_Sample_ID'))
p2 = plot_data %>% ggplot(aes(Mean, color=Diagnosis, fill=Diagnosis)) + geom_density(alpha=0.3) +
                   theme_minimal() + theme(legend.position='bottom') + 
                   ggtitle('Mean expression density by Sample')

grid.arrange(p1, p2, nrow=1)

rm(GO_annotations, plot_data, p1, p2)
```
<br>

### ASD vs CTL

In general there doesn't seem to be a lot of variance in mean expression between autism and control samples by gene. 
<!-- Biggest outliers are ENSG00000173110, ENSG00000234449 and ENSG00000143858. -->
```{r asd_vs_ctl, warning=FALSE}
plot_data = data.frame('ID'=rownames(datExpr),
                       'ASD'=rowMeans(datExpr[,datMeta$Diagnosis=='ASD']),
                       'CTL'=rowMeans(datExpr[,datMeta$Diagnosis!='ASD']))

plot_data %>% ggplot(aes(ASD,CTL)) + geom_point(alpha=0.1, color='#0099cc') + 
         geom_abline(color='gray') + ggtitle('Mean expression ASD vs CTL') + theme_minimal()
```

#### Grouping genes and samples by Diagnosis

- There doesn't seem to be a noticeable difference between mean expression by gene between Diagnosis groups

- Samples with autism tend to have a wider dispersion of mean expression with higher values than the control group (as we had already seen above)

```{r, fig.width=10}
plot_data = rbind(data.frame('Mean'=rowMeans(datExpr[,datMeta$Diagnosis=='ASD']), 'Diagnosis'='ASD'),
                  data.frame('Mean'=rowMeans(datExpr[,datMeta$Diagnosis!='ASD']), 'Diagnosis'='CTL')) %>%
            mutate('Diagnosis'=factor(Diagnosis, levels=c('CTL','ASD')))
p1 = ggplotly(plot_data %>% ggplot(aes(Mean, color=Diagnosis, fill=Diagnosis)) + 
              geom_density(alpha=0.3) + scale_x_log10() + theme_minimal())

plot_data = rbind(data.frame('Mean'=colMeans(datExpr[,datMeta$Diagnosis=='ASD']), 'Diagnosis'='ASD'),
                  data.frame('Mean'=colMeans(datExpr[,datMeta$Diagnosis!='ASD']), 'Diagnosis'='CTL')) %>%
            mutate('Diagnosis'=factor(Diagnosis, levels=c('CTL','ASD')))
p2 = ggplotly(plot_data %>% ggplot(aes(Mean, color=Diagnosis, fill=Diagnosis)) + 
              geom_density(alpha=0.3) + theme_minimal() +
              ggtitle('Mean expression by Gene (left) and by Sample (right) grouped by Diagnosis'))

subplot(p1, p2, nrows=1)

rm(p1, p2, plot_data)
```
<br><br>

---

## Visualisations
<br>

### Samples

#### PCA

The first principal component seems to separate relatively well the two diagnosis, although not as good as when it was with all the brain regions
```{r pca_samples, warning=FALSE, message=FALSE}
pca = datExpr %>% t %>% prcomp

plot_data = data.frame('ID'=colnames(datExpr), 'PC1' = pca$x[,1], 'PC2' = pca$x[,2]) %>% 
            mutate('ID'=substring(ID,2)) %>% left_join(datMeta, by=c('ID'='Dissected_Sample_ID')) %>% 
            dplyr::select('ID','PC1','PC2','Diagnosis') %>% 
            mutate('Diagnosis'=factor(Diagnosis, levels=c('CTL','ASD')))

plot_data %>% ggplot(aes(PC1, PC2, color=Diagnosis)) + geom_point() + theme_minimal() + ggtitle('PCA') +
              xlab(paste0('PC1 (',round(100*summary(pca)$importance[2,1],1),'%)')) +
              ylab(paste0('PC2 (',round(100*summary(pca)$importance[2,2],1),'%)'))

rm(pca, plot_data)
```
<br>

#### MDS

Looks exactly the same as the PCA visualisation, just inverting the x axis
```{r mds_samples, message=FALSE, warning=FALSE}
d = datExpr %>% t %>% dist
fit = cmdscale(d, k=2)

plot_data = data.frame('ID'=colnames(datExpr), 'C1'=fit[,1], 'C2'=fit[,2]) %>%
            mutate('ID'=substring(ID,2)) %>% left_join(datMeta, by=c('ID'='Dissected_Sample_ID')) %>% 
            dplyr::select('C1','C2','Diagnosis') %>%
            mutate('Diagnosis'=factor(Diagnosis, levels=c('CTL','ASD')))

plot_data %>% ggplot(aes(C1, C2, color=Diagnosis)) + geom_point() + theme_minimal() + ggtitle('MDS')

rm(d, fit, plot_data)
```
<br>

#### t-SNE

Higher perplexities seem to capture the difference between diagnosis better. You can separate the diagonsis perfectly with a line using the highest perplexity 

```{r tsne_samples, warning=FALSE, message=FALSE, fig.width=10, fig.height=10}
#perplexities = c(2,5,10,25)
perplexities = c(1,2,3,4)
ps = list()

for(i in 1:length(perplexities)){
  set.seed(123)
  tsne = datExpr %>% t %>% Rtsne(perplexity=perplexities[i])
  plot_data = data.frame('ID'=colnames(datExpr), 'C1'=tsne$Y[,1], 'C2'=tsne$Y[,2]) %>%
              mutate('ID'=substring(ID,2)) %>% left_join(datMeta, by=c('ID'='Dissected_Sample_ID')) %>%
              dplyr::select('C1','C2','Diagnosis','Subject_ID') %>%
              mutate('Diagnosis'=factor(Diagnosis, levels=c('CTL','ASD')))
  ps[[i]] = plot_data %>% ggplot(aes(C1, C2, color=Diagnosis)) + geom_point() + theme_minimal() +
            ggtitle(paste0('Perplexity=',perplexities[i])) + theme(legend.position='none')
}

grid.arrange(grobs=ps, nrow=2)

rm(ps, perplexities, tsne, i)
```

### Genes

#### PCA

- First Principal Component explains over 99% of the total variance

- There's a really strong correlation between the mean expression of a gene and the 1st principal component

```{r pca_genes, fig.width=10, warning=FALSE, message=FALSE}
pca = datExpr %>% prcomp

plot_data = data.frame( 'PC1' = pca$x[,1], 'PC2' = pca$x[,2], 'MeanExpr'=rowMeans(datExpr))

plot_data %>% ggplot(aes(PC1, PC2, color=MeanExpr)) + geom_point(alpha=0.3) + theme_minimal() + 
              scale_color_viridis() + ggtitle('PCA') +
              xlab(paste0('PC1 (',round(100*summary(pca)$importance[2,1],1),'%)')) +
              ylab(paste0('PC2 (',round(100*summary(pca)$importance[2,2],1),'%)'))

rm(pca, plot_data)
```
<br>

#### t-SNE

Higher perplexities capture a cleaner visualisation of the data ordered by mean expression, in a similar (although not as linear) way to PCA
```{r tsne_genes, warning=FALSE, message=FALSE, fig.width=10, fig.height=7}
perplexities = c(1,2,5,10,50,100)
ps = list()

for(i in 1:length(perplexities)){
  tsne = read.csv(paste0('./../Visualisations/tsne_perplexity_',perplexities[i],'.csv'))
  plot_data = data.frame('C1'=tsne[,1], 'C2'=tsne[,2], 'MeanExpr'=rowMeans(datExpr))
  ps[[i]] = plot_data %>% ggplot(aes(C1, C2, color=MeanExpr)) + geom_point(alpha=0.5) + theme_minimal() +
            scale_color_viridis() + ggtitle(paste0('Perplexity = ',perplexities[i])) + theme(legend.position='none')
}

grid.arrange(grobs=ps, nrow=2)

rm(perplexities, ps, tsne, i)
```

<br>

### Differential Expression Analysis

- Only **24** genes (~0.15%) are significant using a threshold of 0.05 for the adjusted p-value and a without a log Fold Change threshold (keeping the null hypothesis $H_0: lfc=0$)


```{r}
cat(paste0(sum(is.na(DE_info$padj)), ' (~', round(100*(sum(is.na(DE_info$padj))/nrow(DE_info))),
           '%) of the p-values couldn\'t be calculated'))

table(DE_info$padj<0.05, useNA='ifany')
```

```{r logFC_vs_adjPval, fig.width=10}
p = DE_info %>% ggplot(aes(log2FoldChange, padj, color=significant)) + geom_point(alpha=0.2) + 
    scale_y_sqrt() + xlab('log2 Fold Change') + ylab('Adjusted p-value') + theme_minimal()
ggExtra::ggMarginal(p, type = 'density', color='gray', fill='gray', size=10)

rm(p)
```

- There is a clear negative relation between lfc and mean expression, for both differentially expressed and not differentially expressed groups of genes

- The relation is strongest for genes with low levels of expression

- The p-value of the genes with the lowest levels of expression couldn't be calculated. DESeq2 does this with genes where it considers the level of expression is not high enough to get robust results

```{r logFC_vs_meanExpr, fig.width=10, warning=FALSE}
plot_data = data.frame('ID'=rownames(datExpr), 'meanExpr'=rowMeans(datExpr)) %>% left_join(DE_info, by='ID') %>%
            mutate('statisticallySignificant' = ifelse(is.na(padj),NA, ifelse(padj<0.05, TRUE, FALSE)),
                   'alpha' = ifelse(padj>0.05 | is.na(padj), 0.1, 0.5))

plot_data %>% ggplot(aes(meanExpr, abs(log2FoldChange), color=statisticallySignificant)) + 
              geom_point(alpha = plot_data$alpha) + geom_smooth(method='lm') + 
              theme_minimal() + scale_y_sqrt() + theme(legend.position = 'bottom') +
              xlab('Mean Expression') + ylab('abs(lfc)') + ggtitle('Log fold change by level of expression')
```

- When filtering for differential expression, the points separate into two clouds depending on whether they are over or underexpressed

- The top cloud corresponds to the over expressed genes and the bottom to the under expressed ones
```{r pca_DE_genes}
datExpr_DE = datExpr[DE_info$significant,]

pca = datExpr_DE %>% prcomp

plot_data = cbind(data.frame('PC1'=pca$x[,1], 'PC2'=pca$x[,2]), DE_info[DE_info$significant==TRUE,])

pos_zero = -min(plot_data$log2FoldChange)/(max(plot_data$log2FoldChange)-min(plot_data$log2FoldChange))
p = plot_data %>% ggplot(aes(PC1, PC2, color=log2FoldChange)) + geom_point(alpha=0.5) +
                  scale_color_gradientn(colours=c('#F8766D','#faa49e','white','#00BFC4','#009499'), 
                                        values=c(0, pos_zero-0.05, pos_zero, pos_zero+0.05, 1)) +
                  theme_minimal() + ggtitle('
PCA of differentially expressed genes') + # This is on purpose, PDF doesn't save well without this white space (?)
                  xlab(paste0('PC1 (',round(100*summary(pca)$importance[2,1],1),'%)')) +
                  ylab(paste0('PC2 (',round(100*summary(pca)$importance[2,2],1),'%)')) + theme(legend.position = 'bottom')
ggExtra::ggMarginal(p, type='density', color='gray', fill='gray', size=10)

rm(pos_zero, p)
```

Separating the genes into these two groups: Salmon: under-expressed, aqua: over-expressed

```{r}
plot_data = plot_data %>% mutate('Group'=ifelse(log2FoldChange>0,'overexpressed','underexpressed')) %>%
            mutate('Group' = factor(Group, levels=c('underexpressed','overexpressed')))

rm(pca)
```

List of DE genes

```{r}
# Get genes names
getinfo = c('ensembl_gene_id','external_gene_id')
mart = useMart(biomart='ENSEMBL_MART_ENSEMBL', dataset='hsapiens_gene_ensembl', host='feb2014.archive.ensembl.org')
gene_names = getBM(attributes=getinfo, filters=c('ensembl_gene_id'), values=plot_data$ID, mart=mart) %>% 
             rename(external_gene_id = 'gene_name', ensembl_gene_id = 'ID')

plot_data = plot_data %>% left_join(gene_names, by='ID')

kable(plot_data %>% dplyr::select(ID, gene_name, log2FoldChange, padj, Neuronal, Group) %>%
      arrange(padj))

```

<br><br>

## Effects of modifying the log fold change threshold

Not only we have very few DE genes, but we lose most of them from very low log Fold Change thresholds

```{r changing_lfc_treshold, warning=FALSE, message=FALSE}

#fc_list = seq(1, 1.04, 0.01)
fc_list = c(seq(1,1.01, 0.002), seq(1.01, 1.04, 0.01))

n_genes = nrow(datExpr)

# Calculate PCAs
datExpr_pca_samps = datExpr %>% data.frame %>% t %>% prcomp(scale.=TRUE)
datExpr_pca_genes = datExpr %>% data.frame %>% prcomp(scale.=TRUE)

# Initialice DF to save PCA outputs
pcas_samps = datExpr_pca_samps$x %>% data.frame %>% dplyr::select(PC1:PC2) %>% 
             mutate('ID'=colnames(datExpr), 'fc'=-1, PC1=scale(PC1), PC2=scale(PC2))
pcas_genes = datExpr_pca_genes$x %>% data.frame %>% dplyr::select(PC1:PC2) %>% 
             mutate('ID'=rownames(datExpr), 'fc'=-1, PC1=scale(PC1), PC2=scale(PC2))

pca_samps_old = pcas_samps
pca_genes_old = pcas_genes

for(fc in fc_list){
  
  # Recalculate DE_info with the new threshold (p-values change) an filter DE genes
  DE_genes = results(dds, lfcThreshold=log2(fc), altHypothesis='greaterAbs') %>% data.frame %>%
             mutate('ID'=rownames(.)) %>% filter(padj<0.05)
  
  datExpr_DE = datExpr %>% data.frame %>% filter(rownames(.) %in% DE_genes$ID)
  n_genes = c(n_genes, nrow(DE_genes))
  
  # Calculate PCAs
  datExpr_pca_samps = datExpr_DE %>% t %>% prcomp(scale.=TRUE)
  datExpr_pca_genes = datExpr_DE %>% prcomp(scale.=TRUE)

  # Create new DF entries
  pca_samps_new = datExpr_pca_samps$x %>% data.frame %>% dplyr::select(PC1:PC2) %>% 
                  mutate('ID'=colnames(datExpr), 'fc'=fc, PC1=scale(PC1), PC2=scale(PC2))
  pca_genes_new = datExpr_pca_genes$x %>% data.frame %>% dplyr::select(PC1:PC2) %>% 
                  mutate('ID'=DE_genes$ID, 'fc'=fc, PC1=scale(PC1), PC2=scale(PC2))  
  
  # Change PC sign if necessary
  if(cor(pca_samps_new$PC1, pca_samps_old$PC1)<0) pca_samps_new$PC1 = -pca_samps_new$PC1
  if(cor(pca_samps_new$PC2, pca_samps_old$PC2)<0) pca_samps_new$PC2 = -pca_samps_new$PC2
  if(cor(pca_genes_new[pca_genes_new$ID %in% pca_genes_old$ID,]$PC1,
         pca_genes_old[pca_genes_old$ID %in% pca_genes_new$ID,]$PC1)<0){
    pca_genes_new$PC1 = -pca_genes_new$PC1
  }
  if(cor(pca_genes_new[pca_genes_new$ID %in% pca_genes_old$ID,]$PC2, 
         pca_genes_old[pca_genes_old$ID %in% pca_genes_new$ID,]$PC2 )<0){
    pca_genes_new$PC2 = -pca_genes_new$PC2
  }
  
  pca_samps_old = pca_samps_new
  pca_genes_old = pca_genes_new
  
  # Update DFs
  pcas_samps = rbind(pcas_samps, pca_samps_new)
  pcas_genes = rbind(pcas_genes, pca_genes_new)
  
}

# Add Diagnosis/SFARI score information
pcas_samps = pcas_samps %>% mutate('ID'=substring(ID,2)) %>% 
             left_join(datMeta, by=c('ID'='Dissected_Sample_ID')) %>%
             dplyr::select(ID, PC1, PC2, fc, Diagnosis, Brain_lobe)
# pcas_genes = pcas_genes %>% left_join(SFARI_genes, by='ID') %>% 
#              mutate('score'=as.factor(`gene-score`)) %>%
#              dplyr::select(ID, PC1, PC2, lfc, score)

# Plot change of number of genes
ggplotly(data.frame('lfc'=log2(fc_list), 'n_genes'=n_genes[-1]) %>% ggplot(aes(x=lfc, y=n_genes)) + 
         geom_point() + geom_line() + theme_minimal() + xlab('|Log Fold Change|') + ylab('Remaining Genes') +
         ggtitle('Remaining genes when modifying filtering threshold'))

rm(fc_list, n_genes, fc, pca_samps_new, pca_genes_new, pca_samps_old, pca_genes_old, 
   datExpr_pca_samps, datExpr_pca_genes)
```

<!-- lfc=0 means no filtering at all, the rest of the filterings include on top of the defined lfc, an adjusted p-value lower than 0.05 -->

<br>

### Samples

**Note:** PC values get smaller as Log2 fold change increases, so on each iteration the values were scaled so it would be easier to compare between frames
<br>

#### Coloured by Diagnosis:

- LFC = -1 represents the whole set of genes, without any filtering by differential expression

- Filtering out genes that are not differentially expressed ($H_0:lfc=0$) separates the two diagnosis groups perfectly using the first Principal Component

- Increasing the LFC threshold doesn't seem to improve the separation between groups much

```{r, warning=FALSE}
ggplotly(pcas_samps %>% mutate(abs_lfc=ifelse(fc==-1,-1,round(log2(abs(fc)),3))) %>% 
         ggplot(aes(PC1, PC2, color=Diagnosis)) + geom_point(aes(frame=abs_lfc, ids=ID)) + 
         theme_minimal() + ggtitle('Samples PCA plot modifying filtering threshold'))
```
<br>

### Genes
```{r, warning=FALSE}

if(!'fcSign' %in% colnames(pcas_genes)){
  pcas_genes = pcas_genes %>% left_join(DE_info, by='ID') %>% mutate(fcSign = ifelse(log2FoldChange>0,'Positive','Negative')) 
}

ggplotly(pcas_genes %>% mutate(abs_lfc=ifelse(fc==-1,-1,round(log2(fc),3))) %>% 
         ggplot(aes(PC1, PC2, color=fcSign)) + geom_point(aes(frame=abs_lfc, ids=ID, alpha=0.1)) + 
         theme_minimal() + ggtitle('Genes PCA plot modifying |LFC| filtering threshold'))
```
<br><br>

---

#### Session info

```{r}
sessionInfo()
```
<br><br>
