---
title: 'Top Modules Exploratory Analysis with New SFARI Genes'
output:
  html_document:
    code_folding: 'hide'
---

```{r load_packages, warning=FALSE, message=FALSE}
#setwd('/afs/inf.ed.ac.uk/user/s17/s1725186/Documents/PhD-Models/FirstPUModel/RMarkdowns')

library(tidyverse) ; library(reshape2) ; library(glue) ; library(plotly) ; library(dendextend)
library(RColorBrewer) ; library(viridis) ; require(gridExtra) ; library(GGally)
library(expss)
library(polycor)
library(foreach) ; library(doParallel)
library(knitr)
library(biomaRt)
library(anRichment) ; library(BrainDiseaseCollection)
suppressWarnings(suppressMessages(library(WGCNA)))
```

Load preprocessed dataset (preprocessing code in 19_10_14_data_preprocessing.Rmd) and clustering (pipeline in 19_10_21_WGCNA.Rmd)
```{r load_dataset, message=FALSE, warning=FALSE}

# Gandal dataset
load('./../Data/preprocessed_data.RData')
datExpr = datExpr %>% data.frame
DE_info = DE_info %>% data.frame


# GO Neuronal annotations: regex 'neuron' in GO functional annotations and label the genes that make a match as neuronal
GO_annotations = read.csv('./../Data/genes_GO_annotations.csv')
GO_neuronal = GO_annotations %>% filter(grepl('neuron', go_term)) %>% 
              mutate('ID'=as.character(ensembl_gene_id)) %>% 
              dplyr::select(-ensembl_gene_id) %>% distinct(ID) %>%
              mutate('Neuronal'=1)


# SFARI Genes
SFARI_genes = read_csv('./../../../SFARI/Data/SFARI_genes_01-03-2020_w_ensembl_IDs.csv')
SFARI_genes = SFARI_genes[!duplicated(SFARI_genes$ID) & !is.na(SFARI_genes$ID),]


# Clusterings
clusterings = read_csv('./../Data/clusters.csv')


# Update DE_info with SFARI and Neuronal information
genes_info = DE_info %>% mutate('ID'=rownames(.)) %>% left_join(SFARI_genes, by='ID') %>% 
  mutate(`gene-score`=ifelse(is.na(`gene-score`), 'None', `gene-score`)) %>%
  left_join(GO_neuronal, by='ID') %>% left_join(clusterings, by='ID') %>%
  mutate(Neuronal=ifelse(is.na(Neuronal), 0, Neuronal)) %>%
  mutate(gene.score=ifelse(`gene-score`=='None' & Neuronal==1, 'Neuronal', `gene-score`), 
         significant=padj<0.05 & !is.na(padj))

# Add gene symbol
getinfo = c('ensembl_gene_id','external_gene_id')
mart = useMart(biomart='ENSEMBL_MART_ENSEMBL', dataset='hsapiens_gene_ensembl',
               host='feb2014.archive.ensembl.org') ## Gencode v19
gene_names = getBM(attributes=getinfo, filters=c('ensembl_gene_id'), values=genes_info$ID, mart=mart)

genes_info = genes_info %>% left_join(gene_names, by=c('ID'='ensembl_gene_id'))


clustering_selected = 'DynamicHybrid'
genes_info$Module = genes_info[,clustering_selected]

dataset = read.csv(paste0('./../Data/dataset_', clustering_selected, '.csv'))
dataset$Module = dataset[,clustering_selected]


SFARI_colour_hue = function(r) {
  pal = c('#FF7631','#FFB100','#E8E328','#b3b3b3','#808080','gray','#d9d9d9')[r]
}

rm(DE_info, GO_annotations, clusterings, getinfo, mart, dds)
```

<br>

## Relation to external clinical traits

<br>

### Quantifying module-trait associations

<br>

Using the `hetcor` function, that calculates Pearson, polyserial or polychoric correlations depending on the type of variables involved.

```{r calc_module_trait_associations, message=FALSE, warning=FALSE}
datTraits = datMeta %>% dplyr::select(Diagnosis, Region, Sex, Age, PMI, RNAExtractionBatch) %>%
            dplyr::rename('ExtractionBatch' = RNAExtractionBatch)

# Recalculate MEs with color labels
ME_object = datExpr %>% t %>% moduleEigengenes(colors = genes_info$Module)
MEs = orderMEs(ME_object$eigengenes)

# Calculate correlation between eigengenes and the traits and their p-values
moduleTraitCor = MEs %>% apply(2, function(x) hetcor(x, datTraits)$correlations[1,-1]) %>% t
rownames(moduleTraitCor) = colnames(MEs)
colnames(moduleTraitCor) = colnames(datTraits)
moduleTraitPvalue = corPvalueStudent(moduleTraitCor, nrow(datExpr))

# Create text matrix for the Heatmap
textMatrix = paste0(signif(moduleTraitCor, 2), ' (', signif(moduleTraitPvalue, 1), ')')
dim(textMatrix) = dim(moduleTraitCor)

# In case there are any NAs
if(sum(!complete.cases(moduleTraitCor))>0){
  print(paste0(sum(is.na(moduleTraitCor)),' correlation(s) could not be calculated')) 
}

rm(ME_object)
```

<!-- **Note:** The correlation between Module #F464E5 and Diagonsis is the one that cannot be calculated, weirdly enough, the thing that causes the error is that the initial correlation is too high, so it would be a very bad thing to lose this module because of this numerical error. I’m going to fill in its value using the polyserial function, which doesn’t give exactly the same results as the hetcor() function, but it’s quite similar. -->

<!-- ```{r} -->
<!-- # Calculate the correlation tha failed with hetcor() -->
<!-- moduleTraitCor['ME#F464E5','Diagnosis'] = polyserial(MEs[,'ME#F464E5'], datTraits$Diagnosis) -->
<!-- ``` -->

I'm going to select all the modules that have an absolute correlation higher than 0.9 with Diagnosis to study them
```{r plot_heatmap, fig.width=10, fig.height=10, warning=FALSE, message=FALSE}

# Sort moduleTraitCor by Diagnosis
moduleTraitCor = moduleTraitCor[order(moduleTraitCor[,1], decreasing=TRUE),]
moduleTraitPvalue = moduleTraitPvalue[order(moduleTraitCor[,1], decreasing=TRUE),]

# Create text matrix for the Heatmap
textMatrix = paste0(signif(moduleTraitCor, 2), ' (', signif(moduleTraitPvalue, 1), ')')
dim(textMatrix) = dim(moduleTraitCor)


labeledHeatmap(Matrix = moduleTraitCor, xLabels = names(datTraits), yLabels =  gsub('ME','',rownames(moduleTraitCor)), 
               yColorWidth=0, colors = brewer.pal(11,'PiYG'), bg.lab.y = gsub('ME','',rownames(moduleTraitCor)),
               textMatrix = textMatrix, setStdMargins = FALSE, cex.text = 0.8, cex.lab.y = 0.75, zlim = c(-1,1),
               main = paste('Module-Trait relationships'))

diagnosis_cor = data.frame('Module' = gsub('ME','',rownames(moduleTraitCor)),
                           'MTcor' = moduleTraitCor[,'Diagnosis'],
                           'MTpval' = moduleTraitPvalue[,'Diagnosis'])

genes_info = genes_info %>% left_join(diagnosis_cor, by='Module')

rm(moduleTraitPvalue, datTraits, textMatrix, diagnosis_cor)
```
<br>

## Studying the modules with the highest absolute correlation to Diagnosis

<br>

The modules consist mainly of points with very high (absolute) values in PC2 (which we know is related to lfc), so this result is consistent with the high correlation between Module and Diagnosis, although some of the points with the highest PC2 values do not belong to these top modules

```{r warning=FALSE}
top_modules = gsub('ME','',rownames(moduleTraitCor)[abs(moduleTraitCor[,'Diagnosis'])>0.9])

cat(paste0('Top modules selected: ', paste(top_modules, collapse=', '),'\n'))

pca = datExpr %>% prcomp

plot_data = data.frame('ID'=rownames(datExpr), 'PC1' = pca$x[,1], 'PC2' = pca$x[,2]) %>%
            left_join(dataset, by='ID') %>% left_join(genes_info %>% dplyr::select(ID, external_gene_id), by='ID') %>%
            dplyr::select(ID, external_gene_id, PC1, PC2, Module, gene.score) %>%
            mutate(ImportantModules = ifelse(Module %in% top_modules, as.character(Module), 'Others')) %>%
            mutate(color = ifelse(ImportantModules=='Others','gray',ImportantModules),
                   alpha = ifelse(ImportantModules=='Others', 0.2, 0.4),
                   gene_id = paste0(ID, ' (', external_gene_id, ')'))

table(plot_data$ImportantModules)

ggplotly(plot_data %>% ggplot(aes(PC1, PC2, color=ImportantModules)) + 
         geom_point(alpha=plot_data$alpha, color=plot_data$color, aes(ID=gene_id)) + theme_minimal() + 
           ggtitle('Modules with strongest relation to Diagnosis'))

rm(pca)
```

<br>

---

<br>

### Module Membership vs Gene Significance

<br>

```{r, warning=FALSE}

create_plot = function(module){
  
  plot_data = dataset %>% dplyr::select(ID, paste0('MM.',gsub('#','',module)), GS, gene.score) %>% filter(dataset$Module==module)
  colnames(plot_data)[2] = 'Module'
  
  SFARI_colors = as.numeric(names(table(as.character(plot_data$gene.score)[plot_data$gene.score!='None'])))
  
  p = ggplotly(plot_data %>% ggplot(aes(Module, GS, color=gene.score)) + geom_point(alpha=0.5, aes(ID=ID)) + ylab('Gene Significance') +
               scale_color_manual(values=SFARI_colour_hue(r=c(SFARI_colors,8))) + theme_minimal() + xlab('Module Membership') +
               ggtitle(paste0('Module ', module,'  (MTcor = ', round(moduleTraitCor[paste0('ME',module),1],2),')')))
  
  return(p)
}

create_plot(top_modules[1])
create_plot(top_modules[2])
create_plot(top_modules[3])
create_plot(top_modules[4])

rm(create_plot)
```

<br>

---

<br>

### SFARI Genes

<br>

List of top SFARI Genes in top modules ordered by SFARI score and Gene Significance
```{r, warning=FALSE}

table_data = dataset %>% left_join(genes_info %>% dplyr::select(ID, external_gene_id), by='ID') %>%
             dplyr::select(ID, external_gene_id, GS, gene.score, Module) %>% arrange(gene.score, desc(abs(GS))) %>%
             dplyr::rename('Ensembl ID'=ID, 'Gene Symbol'=external_gene_id, 
                    'SFARI score'=gene.score, 'Gene Significance'=GS)

kable(table_data %>% filter(Module == top_modules[1] & `SFARI score` %in% c(1,2,3)) %>% dplyr::select(-Module),
      caption=paste0('Top SFARI Genes for Module ', top_modules[1]))

kable(table_data %>% filter(Module == top_modules[2] & `SFARI score` %in% c(1,2,3)) %>% dplyr::select(-Module),
      caption=paste0('Top SFARI Genes for Module ', top_modules[2]))

kable(table_data %>% filter(Module == top_modules[3] & `SFARI score` %in% c(1,2,3)) %>% dplyr::select(-Module),
      caption=paste0('Top SFARI Genes for Module ', top_modules[3]))

kable(table_data %>% filter(Module == top_modules[4] & `SFARI score` %in% c(1,2,3)) %>% dplyr::select(-Module),
      caption=paste0('Top SFARI Genes for Module ', top_modules[4]))

```

Modules with the strongest module-diagnosis correlation should have the highest percentage of SFARI Genes, but this doesn't seem to be the case (even the opposite may be true)

```{r, warning=FALSE, message=FALSE}
plot_data = dataset %>% mutate('hasSFARIscore' = gene.score!='None') %>% 
            group_by(Module, MTcor, hasSFARIscore) %>% summarise(p=n()) %>% 
            left_join(dataset %>% group_by(Module) %>% summarise(n=n()), by='Module') %>% 
            mutate(p=round(p/n*100,2)) 

for(i in 1:nrow(plot_data)){
  this_row = plot_data[i,]
  if(this_row$hasSFARIscore==FALSE & this_row$p==100){
    new_row = this_row
    new_row$hasSFARIscore = TRUE
    new_row$p = 0
    plot_data = plot_data %>% rbind(new_row)
  }
}

plot_data = plot_data %>% filter(hasSFARIscore==TRUE)

ggplotly(plot_data %>% ggplot(aes(MTcor, p, size=n)) + geom_smooth(color='gray', se=FALSE) +
         geom_point(color=plot_data$Module, alpha=0.5, aes(id=Module)) + geom_hline(yintercept=mean(plot_data$p), color='gray') +
         xlab('Module-Diagnosis correlation') + ylab('% of SFARI genes') +
         theme_minimal() + theme(legend.position = 'none'))

rm(i, this_row, new_row, plot_data)
```

Breaking the SFARI genes by score
```{r, warning=FALSE, message=FALSE, fig.width=10}
scores = c(1,2,3,'None')

plot_data = dataset %>% group_by(Module, MTcor, gene.score) %>% summarise(n=n()) %>% 
            left_join(dataset %>% group_by(Module) %>% summarise(N=n()), by='Module') %>% 
            mutate(p=round(n/N*100,2), gene.score = as.character(gene.score))

for(i in 1:nrow(plot_data)){
  this_row = plot_data[i,]
  if(sum(plot_data$Module == this_row$Module)<7){
    missing_scores = which(! scores %in% plot_data$gene.score[plot_data$Module == this_row$Module])
    for(s in missing_scores){
      new_row = this_row
      new_row$gene.score = as.character(s)
      new_row$n = 0
      new_row$p = 0
      plot_data = plot_data %>% rbind(new_row) 
    }
  }
}

plot_data = plot_data %>% filter(gene.score != 'None')

plot_function = function(i){
  i = 2*i-1
  plot_list = list()
  for(j in 1:2){
    plot_list[[j]] = ggplotly(plot_data %>% filter(gene.score==scores[i+j-1]) %>% ggplot(aes(MTcor, p, size=n)) + 
                geom_smooth(color='gray', se=FALSE) +
                geom_point(color=plot_data$Module[plot_data$gene.score==scores[i+j-1]], alpha=0.5, aes(id=Module)) +
                geom_hline(yintercept=mean(plot_data$p[plot_data$gene.score==scores[i+j-1]]), color='gray') +
                xlab('Module-Diagnosis correlation') + ylab('% of SFARI genes') +
                theme_minimal() + theme(legend.position = 'none'))
  }
  p = subplot(plot_list, nrows=1) %>% layout(annotations = list(
    list(x = 0.2 , y = 1.05, text = paste0('SFARI score ', scores[i]), showarrow = F, xref='paper', yref='paper'),
    list(x = 0.8 , y = 1.05, text = paste0('SFARI score ', scores[i+1]), showarrow = F, xref='paper', yref='paper')))
  
  return(p)
}

plot_function(1)
plot_function(2)

rm(i, s, this_row, new_row, plot_function)
```

<br>

---

<br>

### Module Eigengenes

<br>

Since these modules have the strongest relation to autism, this pattern should be reflected in their model eigengenes, having two different behaviours for the samples corresponding to autism and the ones corresponding to control.

In both cases, the Eigengenes separate the behaviour between autism and control samples very clearly!
```{r, fig.width=10, fig.height=10}

plot_EGs = function(module){

  plot_data = data.frame('ID' = rownames(MEs), 'MEs' = MEs[,paste0('ME',module)], 'Diagnosis' = datMeta$Diagnosis)

  p = plot_data %>% ggplot(aes(Diagnosis, MEs, fill=Diagnosis)) + geom_boxplot() + theme_minimal() + theme(legend.position='none') +
                    ggtitle(paste0('Module ', module, '  (MTcor=',round(moduleTraitCor[paste0('ME',module),1],2),')'))
  return(p)
}

p1 = plot_EGs(top_modules[1])
p2 = plot_EGs(top_modules[2])
p3 = plot_EGs(top_modules[3])
p4 = plot_EGs(top_modules[4])

grid.arrange(p1, p2, p3, p4, nrow=2)

rm(plot_EGs, p1, p2, p3, p4)
```

<br>

---

<br>

### Identifying important genes

<br>

Selecting the modules with the highest correlation to Diagnosis, and, from them, the genes with the highest module membership-(absolute) gene significance

*Ordered by $\frac{MM+|GS|}{2}$

There aren't that many SFARI genes in the top genes of the modules and not a single one belonging to scores 1 and 2
```{r select_important_genes, warning=FALSE, message=FALSE}

create_table = function(module){
  top_genes = dataset %>% left_join(genes_info %>% dplyr::select(ID, external_gene_id), by='ID') %>% 
              dplyr::select(ID, external_gene_id, paste0('MM.',gsub('#','',module)), GS, gene.score) %>%
              filter(dataset$Module==module) %>% dplyr::rename('MM' = paste0('MM.',gsub('#','',module))) %>% 
              mutate(importance = (MM+abs(GS))/2) %>% arrange(by=-importance) %>% top_n(20)
  return(top_genes)
}

top_genes = list()
for(i in 1:length(top_modules)) top_genes[[i]] = create_table(top_modules[i])

kable(top_genes[[1]], caption=paste0('Top 10 genes for module ', top_modules[1], '  (MTcor = ',
                                  round(moduleTraitCor[paste0('ME',top_modules[1]),1],2),')'))

kable(top_genes[[2]], caption=paste0('Top 10 genes for module ', top_modules[2], '  (MTcor = ',
                                  round(moduleTraitCor[paste0('ME',top_modules[2]),1],2),')'))

kable(top_genes[[3]], caption=paste0('Top 10 genes for module ', top_modules[3], '  (MTcor = ',
                                  round(moduleTraitCor[paste0('ME',top_modules[3]),1],2),')'))

kable(top_genes[[4]], caption=paste0('Top 10 genes for module ', top_modules[4], '  (MTcor = ',
                                  round(moduleTraitCor[paste0('ME',top_modules[4]),1],2),')'))
rm(create_table)
```

```{r}
pca = datExpr %>% prcomp

ids = c()
for(tg in top_genes) ids = c(ids, tg$ID)

plot_data = data.frame('ID'=rownames(datExpr), 'PC1' = pca$x[,1], 'PC2' = pca$x[,2]) %>%
            left_join(dataset, by='ID') %>% dplyr::select(ID, PC1, PC2, Module, gene.score) %>%
            mutate(color = ifelse(Module %in% top_modules, as.character(Module), 'gray')) %>%
            mutate(alpha = ifelse(color %in% top_modules & 
                                  ID %in% ids, 1, 0.1))

plot_data %>% ggplot(aes(PC1, PC2)) + geom_point(alpha=plot_data$alpha, color=plot_data$color) + 
              theme_minimal() + ggtitle('Important genes identified through WGCNA')
```

Level of expression by Diagnosis for top genes, ordered by importance (defined above)
```{r, warning=FALSE, fig.width=10}
create_plot = function(i){
  
  plot_data = datExpr[rownames(datExpr) %in% top_genes[[i]]$ID,] %>% mutate('ID' = rownames(.)) %>% 
              melt(id.vars='ID') %>% mutate(variable = gsub('X','',variable)) %>%
              left_join(top_genes[[i]], by='ID') %>%
              left_join(datMeta %>% dplyr::select(Dissected_Sample_ID, Diagnosis),
                        by = c('variable'='Dissected_Sample_ID')) %>% arrange(desc(importance))
  
  p = ggplotly(plot_data %>% mutate(external_gene_id=factor(external_gene_id, 
                                    levels=unique(plot_data$external_gene_id), ordered=T)) %>%
               ggplot(aes(external_gene_id, value, fill=Diagnosis)) + geom_boxplot() + theme_minimal() +
                      xlab(paste0('Top genes for module ', top_modules[i], ' (MTcor = ',
                       round(genes_info$MTcor[genes_info$Module==top_modules[i]][1],2), ')')) + ylab('Level of Expression') +
                      theme(axis.text.x = element_text(angle = 90, hjust = 1)))
  return(p)
  
}

create_plot(1)
create_plot(2)
create_plot(3)
create_plot(4)


rm(create_plot)
```

<br>

---

<br>

## Enrichment Analysis

<br>

Using the package [anRichment](https://peterlangfelder.com/2018/11/25/functional-enrichment-analysis-via-r-package-anrichment/)

- It was designed by Peter Langfelder explicitly to perform enrichmen analysis on WGCNA's modules in brain-related experiments (mainly Huntington's Disease)

- It has packages with brain annotations:

  - [BrainDiseaseCollection:](https://horvath.genetics.ucla.edu/html/CoexpressionNetwork/GeneAnnotation/BrainDiseaseCollection/) A Brain Disease Gene Set Collection for anRichment
  
  - MillerAIBSCollection: (included in anRichment) Contains gene sets collected by Jeremy A. Miller at AIBS of various cell type and brain region marker sets, gene sets collected from expression studies of developing brain, as well as a collection of transcription factor (TF) targets from the original ChEA study

- The tutorial says it's an experimental package

- It's not on CRAN nor in Bioconductor

```{r, warning=FALSE}
# Prepare dataset

# Create dataset with top modules membership and removing the genes without an assigned module
EA_dataset = data.frame('ensembl_gene_id' = genes_info$ID,
                        module = ifelse(genes_info$Module %in% top_modules, genes_info$Module, 'other')) %>%
             filter(genes_info$Module!='gray')

# Assign Entrez Gene Id to each gene
getinfo = c('ensembl_gene_id','entrezgene')
mart = useMart(biomart='ENSEMBL_MART_ENSEMBL', dataset='hsapiens_gene_ensembl', host='feb2014.archive.ensembl.org')
biomart_output = getBM(attributes=getinfo, filters=c('ensembl_gene_id'), values=EA_dataset$ensembl_gene_id, mart=mart)

EA_dataset = EA_dataset %>% left_join(biomart_output, by='ensembl_gene_id')

for(tm in top_modules){
  cat(paste0('\n',sum(EA_dataset$module==tm & is.na(EA_dataset$entrezgene)), ' genes from top module ',
             tm, ' don\'t have an Entrez Gene ID')) 
}

rm(getinfo, mart, biomart_output, tm)
```

```{r enrichment_analysis, warning=FALSE}

# Manual: https://horvath.genetics.ucla.edu/html/CoexpressionNetwork/GeneAnnotation/Tutorials/anRichment-Tutorial1.pdf

collectGarbage()

# EA_dataset = rbind(EA_dataset[EA_dataset$module!='other',], EA_dataset[EA_dataset$module=='other',][sample(sum(EA_dataset$module=='other'), 1000),])

# Prepare datasets
GO_col = buildGOcollection(organism = 'human', verbose = 0)
internal_col = internalCollection(organism = 'human')
MillerAIBS_col = MillerAIBSCollection(organism = 'human')
BrainDisease_col = BrainDiseaseCollection(organism = 'human')
combined_col = mergeCollections(GO_col, internal_col, MillerAIBS_col, BrainDisease_col)

# Print collections used
cat('Using collections: ')
knownGroups(combined_col, sortBy = 'size')

# Perform Enrichment Analysis
enrichment = enrichmentAnalysis(classLabels = EA_dataset$module, identifiers = EA_dataset$entrezgene,
                                refCollection = combined_col, #useBackground = 'given', 
                                threshold = 1e-4, thresholdType = 'Bonferroni',
                                getOverlapEntrez = FALSE, getOverlapSymbols = TRUE)

```

<br>

### Results

<br>

```{r}

kable(enrichment$enrichmentTable %>% filter(class==top_modules[1]) %>% 
      dplyr::select(dataSetID, shortDataSetName, inGroups, Bonferroni, FDR, enrichmentRatio,
                    effectiveClassSize, effectiveSetSize, nCommonGenes) %>%
      arrange(Bonferroni, desc(enrichmentRatio)),
      caption = paste0('Enriched terms for module ', top_modules[1], ' (MTcor = ',
                       round(genes_info$MTcor[genes_info$Module==top_modules[1]][1],4), ')'))


kable(enrichment$enrichmentTable %>% filter(class==top_modules[2]) %>% 
      dplyr::select(dataSetID, shortDataSetName, inGroups, Bonferroni, FDR, enrichmentRatio,
                    effectiveClassSize, effectiveSetSize, nCommonGenes) %>%
      arrange(Bonferroni, enrichmentRatio),
      caption = paste0('Enriched terms for module ', top_modules[2], ' (MTcor = ',
                       round(genes_info$MTcor[genes_info$Module==top_modules[2]][1],4), ')'))


kable(enrichment$enrichmentTable %>% filter(class==top_modules[3]) %>% 
      dplyr::select(dataSetID, shortDataSetName, inGroups, Bonferroni, FDR, enrichmentRatio,
                    effectiveClassSize, effectiveSetSize, nCommonGenes) %>%
      arrange(Bonferroni, desc(enrichmentRatio)),
      caption = paste0('Enriched terms for module ', top_modules[3], ' (MTcor = ',
                       round(genes_info$MTcor[genes_info$Module==top_modules[3]][1],4), ')'))


kable(enrichment$enrichmentTable %>% filter(class==top_modules[4]) %>% 
      dplyr::select(dataSetID, shortDataSetName, inGroups, Bonferroni, FDR, enrichmentRatio,
                    effectiveClassSize, effectiveSetSize, nCommonGenes) %>%
      arrange(Bonferroni, desc(enrichmentRatio)),
      caption = paste0('Enriched terms for module ', top_modules[4], ' (MTcor = ',
                       round(genes_info$MTcor[genes_info$Module==top_modules[4]][1],4), ')'))

```

Save Enrichment Analysis results
```{r}
save(enrichment, file='./../Data/enrichmentAnalysis.RData')
#load('./../Data/enrichmentAnalysis.RData')
```


<br>

---

<br>

## Enrichment Analysis for Modules with High Percentage of SFARI Genes

<br>

**Option 1:** Select the modules with the highest percentage of SFARI Genes

- This is an easy approach but it doesn't take into account the size of the modules (the bigger the module, the more robust the result)

- Perhaps an approach that incorporates the cluster size into the selection criteria would be better

```{r, warning=FALSE, fig.width=10}
# Prepare dataset
perc_SFARI_genes = dataset %>% mutate('hasSFARIscore' = gene.score!='None') %>% 
                   group_by(Module, MTcor, hasSFARIscore) %>% summarise(p=n()) %>% 
                   left_join(dataset %>% group_by(Module) %>% summarise(n=n()), by='Module') %>% 
                   mutate(p=round(p/n*100,2)) %>% filter(hasSFARIscore & Module != 'gray') %>% arrange(desc(p))

perc_SFARI_genes %>% ggplot(aes(reorder(Module, -p), p)) + geom_bar(stat='identity', fill = perc_SFARI_genes$Module) + 
                     xlab('Module') + ylab('% of SFARI Genes') + theme_minimal() + theme(axis.text.x = element_text(angle = 90, hjust = 1))

keep_modules = perc_SFARI_genes$Module[1:2]

cat(paste0('Keeping top 2 modules: ', paste(keep_modules, collapse = ', ')))
```

**Option 2:** Assume independence between SFARI Genes and modules and calculate the probability distribution of 




```{r, warning=FALSE}
# Create dataset with top modules membership and removing the genes without an assigned module
EA_dataset = data.frame('ensembl_gene_id' = genes_info$ID,
                        module = ifelse(genes_info$Module %in% keep_modules, genes_info$Module, 'other')) %>%
                        filter(genes_info$Module!='gray')

# Assign Entrez Gene Id to each gene
getinfo = c('ensembl_gene_id','entrezgene')
mart = useMart(biomart='ENSEMBL_MART_ENSEMBL', dataset='hsapiens_gene_ensembl', host='feb2014.archive.ensembl.org')
biomart_output = getBM(attributes=getinfo, filters=c('ensembl_gene_id'), values=EA_dataset$ensembl_gene_id, mart=mart)

EA_dataset = EA_dataset %>% left_join(biomart_output, by='ensembl_gene_id')

for(tm in keep_modules){
  cat(paste0('\n',sum(EA_dataset$module==tm & is.na(EA_dataset$entrezgene)), ' genes from top module ',
             tm, ' don\'t have an Entrez Gene ID')) 
}

rm(getinfo, mart, biomart_output, tm)
```

```{r enrichment_analysis_SFARI, warning=FALSE}

# Manual: https://horvath.genetics.ucla.edu/html/CoexpressionNetwork/GeneAnnotation/Tutorials/anRichment-Tutorial1.pdf

collectGarbage()

# EA_dataset = rbind(EA_dataset[EA_dataset$module!='other',], EA_dataset[EA_dataset$module=='other',][sample(sum(EA_dataset$module=='other'), 1000),])

# Prepare datasets
GO_col = buildGOcollection(organism = 'human', verbose = 0)
internal_col = internalCollection(organism = 'human')
MillerAIBS_col = MillerAIBSCollection(organism = 'human')
BrainDisease_col = BrainDiseaseCollection(organism = 'human')
combined_col = mergeCollections(GO_col, internal_col, MillerAIBS_col, BrainDisease_col)

# Print collections used
cat('Using collections: ')
knownGroups(combined_col, sortBy = 'size')

# Perform Enrichment Analysis
enrichment = enrichmentAnalysis(classLabels = EA_dataset$module, identifiers = EA_dataset$entrezgene,
                                refCollection = combined_col, #useBackground = 'given', 
                                threshold = 1e-4, thresholdType = 'Bonferroni',
                                getOverlapEntrez = FALSE, getOverlapSymbols = TRUE)

```

<br>

### Results

<br>

```{r}

kable(enrichment$enrichmentTable %>% filter(class==keep_modules[1]) %>% 
      dplyr::select(dataSetID, shortDataSetName, inGroups, Bonferroni, FDR, enrichmentRatio,
                    effectiveClassSize, effectiveSetSize, nCommonGenes) %>%
      arrange(Bonferroni, desc(enrichmentRatio)),
      caption = paste0('Enriched terms for module ', keep_modules[1], ' (SFARI Genes = ',
                       round(perc_SFARI_genes$p[perc_SFARI_genes$Module==keep_modules[1]][1],4), '%)'))


kable(enrichment$enrichmentTable %>% filter(class==keep_modules[2]) %>% 
      dplyr::select(dataSetID, shortDataSetName, inGroups, Bonferroni, FDR, enrichmentRatio,
                    effectiveClassSize, effectiveSetSize, nCommonGenes) %>%
      arrange(Bonferroni, enrichmentRatio),
      caption = paste0('Enriched terms for module ', keep_modules[2], ' (SFARI Genes = ',
                       round(perc_SFARI_genes$p[perc_SFARI_genes$Module==keep_modules[2]][1],4), '%)'))

```

Save Enrichment Analysis results
```{r}
save(enrichment, file='./../Data/enrichmentAnalysis_SFARI.RData')
#load('./../Data/enrichmentAnalysis.RData')
```

<br><br>

#### Session info

```{r print_session_info}
sessionInfo()
```
<br><br>
