---
title: 'Enrichment Analysis of Modules with highest Enrichment in SFARI Genes'
output:
  html_document:
    code_folding: 'hide'
---

```{r load_packages, warning=FALSE, message=FALSE}
#setwd('/afs/inf.ed.ac.uk/user/s17/s1725186/Documents/PhD-Models/FirstPUModel/RMarkdowns')

library(tidyverse) ; library(reshape2) ; library(glue) ; library(plotly) ; library(dendextend)
library(RColorBrewer) ; library(viridis) ; require(gridExtra) ; library(GGally)
library(expss)
library(knitr)
library(anRichment) #; library(BrainDiseaseCollection)
```


Load preprocessed dataset (preprocessing code in 19_10_14_data_preprocessing.Rmd) and clustering (pipeline in 19_10_21_WGCNA.Rmd)
```{r load_dataset, message=FALSE, warning=FALSE}

# Gandal dataset
load('./../Data/preprocessed_data.RData')
datExpr = datExpr %>% data.frame
DE_info = DE_info %>% data.frame


# GO Neuronal annotations: regex 'neuron' in GO functional annotations and label the genes that make a match as neuronal
GO_annotations = read.csv('./../Data/genes_GO_annotations.csv')
GO_neuronal = GO_annotations %>% filter(grepl('neuron', go_term)) %>% 
              mutate('ID'=as.character(ensembl_gene_id)) %>% 
              dplyr::select(-ensembl_gene_id) %>% distinct(ID) %>%
              mutate('Neuronal'=1)


# SFARI Genes
SFARI_genes = read_csv('./../../../SFARI/Data/SFARI_genes_01-03-2020_w_ensembl_IDs.csv')
SFARI_genes = SFARI_genes[!duplicated(SFARI_genes$ID) & !is.na(SFARI_genes$ID),]


# Clusterings
clusterings = read_csv('./../Data/clusters.csv')


# Update DE_info with SFARI and Neuronal information
genes_info = DE_info %>% mutate('ID'=rownames(.)) %>% left_join(SFARI_genes, by='ID') %>% 
  mutate(`gene-score`=ifelse(is.na(`gene-score`), 'None', `gene-score`)) %>%
  left_join(GO_neuronal, by='ID') %>% left_join(clusterings, by='ID') %>%
  mutate(Neuronal=ifelse(is.na(Neuronal), 0, Neuronal)) %>%
  mutate(gene.score=ifelse(`gene-score`=='None' & Neuronal==1, 'Neuronal', `gene-score`), 
         significant=padj<0.05 & !is.na(padj))


clustering_selected = 'DynamicHybrid'
genes_info$Module = genes_info[,clustering_selected]

dataset = read.csv(paste0('./../Data/dataset_', clustering_selected, '.csv'))
dataset$Module = dataset[,clustering_selected]

# Correct SFARI Scores
dataset$gene.score = genes_info$gene.score

# Enrichment Analysis
load('./../Data/enrichmentAnalysis.RData')


SFARI_colour_hue = function(r) {
  pal = c('#FF7631','#FFB100','#E8E328','#8CC83F','#62CCA6','#59B9C9','#b3b3b3','#808080','gray','#d9d9d9')[r]
}

rm(DE_info, GO_annotations, clusterings, getinfo, mart, dds, GO_neuronal)
```


## Calculate the Enrichment of SFARI Genes in Modules
<br>

This is explained in 20_05_22_WGCNA_top_SFARI_modules.html

```{r warning=FALSE, message=FALSE}
SFARI_genes_by_module = dataset %>% mutate('hasSFARIscore' = !gene.score %in% c('None', 'Neuronal')) %>% 
                        group_by(Module, MTcor, hasSFARIscore) %>% summarise(s=n()) %>% 
                        left_join(dataset %>% group_by(Module) %>% summarise(n=n()), by='Module') %>% 
                        mutate(perc=round(s/n*100,2)) %>% filter(hasSFARIscore & Module != 'gray') %>% arrange(desc(perc)) %>% ungroup

N = sum(SFARI_genes_by_module$n)
S = sum(SFARI_genes_by_module$s)

calc_prob = function(row, log.p, S){
  s = row[4] %>% as.numeric
  n = row[5] %>% as.numeric
  prob = phyper(s, S, N-S, n, log.p = log.p, lower.tail=FALSE)
  return(prob)
}

SFARI_genes_by_module$prob = apply(SFARI_genes_by_module, 1, function(x) calc_prob(x, FALSE, S))
SFARI_genes_by_module$adj_prob = p.adjust(SFARI_genes_by_module$prob, method = 'bonferroni')
SFARI_genes_by_module = SFARI_genes_by_module %>% arrange(prob)

ggplotly(SFARI_genes_by_module %>% ggplot(aes(MTcor, prob, size=n)) + geom_point(color=SFARI_genes_by_module$Module, alpha=0.5, aes(id=Module)) + 
         geom_smooth(color='#cccccc', size = 0.5, se=FALSE) + xlab('Module-Diagnosis Correlation') + ylab('Probability') + 
         ggtitle(paste0('Corr = ', round(cor(SFARI_genes_by_module$MTcor, SFARI_genes_by_module$prob),2), ': Corr[Module-ASD corr<0] = ', 
                        round(cor(SFARI_genes_by_module$MTcor[SFARI_genes_by_module$MTcor<0], SFARI_genes_by_module$prob[SFARI_genes_by_module$MTcor<0]),3),
                        ' Corr[Module-ASD corr>0] = ',
                        round(cor(SFARI_genes_by_module$MTcor[SFARI_genes_by_module$MTcor>=0], SFARI_genes_by_module$prob[SFARI_genes_by_module$MTcor>=0]),2))) +
         theme_minimal() + theme(legend.position = 'none'))

```

<br><br>

### Selecting Top Modules Enriched in SFARI Genes
<br>

We can interpet the probability we obtain as a p-value (and correct it for multiple testing), we can use it as a threshold to identify modules with a significantly high percentage of SFARI genes (adjusted p-value < 0.01)

Using log-scale to help us differentiate between small differences close to zero better

```{r warning=FALSE, message=FALSE}

ggplotly(SFARI_genes_by_module %>% ggplot(aes(MTcor, adj_prob, size=n)) + geom_point(color=SFARI_genes_by_module$Module, alpha=0.5, aes(id=Module)) +
         geom_hline(yintercept = 0.05, color = 'gray', linetype = 'dotted') + xlab('Module-Diagnosis Correlation') + ylab('Corrected p-values') + scale_y_log10() +
         theme_minimal() + theme(legend.position = 'none'))

top_modules = SFARI_genes_by_module$Module[SFARI_genes_by_module$adj_prob <0.05]

cat(paste0('Keeping top ', length(top_modules),' modules: ', paste(top_modules, collapse = ', ')))

rm(N,S,calc_prob)
```

<br>

## Enrichment Analysis

```{r}
file_name = './../Data/enrichmentAnalysis.RData'

if(file.exists(file_name)){
  load(file_name)
} else {
  
  ##############################################################################
  # PREPARE DATASET
  # Create dataset with top modules membership and removing the genes without an assigned module
  EA_dataset = data.frame('ensembl_gene_id' = genes_info$ID, module = genes_info$Module)  %>%
               filter(genes_info$Module!='gray')
  
  # Assign Entrez Gene Id to each gene
  getinfo = c('ensembl_gene_id','entrezgene')
  mart = useMart(biomart='ENSEMBL_MART_ENSEMBL', dataset='hsapiens_gene_ensembl', host='feb2014.archive.ensembl.org')
  biomart_output = getBM(attributes=getinfo, filters=c('ensembl_gene_id'), values=EA_dataset$ensembl_gene_id, mart=mart)
  
  EA_dataset = EA_dataset %>% left_join(biomart_output, by='ensembl_gene_id')
  
  for(tm in top_modules){
    cat(paste0('\n',sum(EA_dataset$module==tm & is.na(EA_dataset$entrezgene)), ' genes from top module ',
               tm, ' don\'t have an Entrez Gene ID')) 
  }
  
  rm(getinfo, mart, biomart_output, tm)
  
  
  ##############################################################################
  # PERFORM ENRICHMENT ANALYSIS
  # Manual: https://horvath.genetics.ucla.edu/html/CoexpressionNetwork/GeneAnnotation/Tutorials/anRichment-Tutorial1.pdf
  
  collectGarbage()
  
  # Prepare datasets
  GO_col = buildGOcollection(organism = 'human', verbose = 0)
  biosys_col = BioSystemsCollection('human')
  internal_col = internalCollection(organism = 'human')
  #MillerAIBS_col = MillerAIBSCollection(organism = 'human') # Information already contained in internalCollection
  #BrainDisease_col = BrainDiseaseCollection(organism = 'human') # Information already contained in internalCollection
  
  combined_col = mergeCollections(GO_col, biosys_col, internal_col)
  
  # Print collections used
  cat('Using collections: ')
  knownGroups(combined_col, sortBy = 'size')
  
  # Perform Enrichment Analysis
  enrichment = enrichmentAnalysis(classLabels = EA_dataset$module, identifiers = EA_dataset$entrezgene,
                                  refCollection = combined_col, #useBackground = 'given', 
                                  threshold = 1e-4, thresholdType = 'Bonferroni',
                                  getOverlapEntrez = FALSE, getOverlapSymbols = TRUE)
  
  # Save enrichment results
  save(enrichment, file=file_name)
  
  rm(GO_col, biosys_col, internal_col, BrainDisease_col, combined_col, EA_dataset)
}

```
<br>

### KEGG Results

```{r, results = 'asis'}

for(i in 1:length(top_modules)){
  df = enrichment$enrichmentTable %>% filter(class==top_modules[i] & grepl('BioSystems.KEGG',dataSetID) & Bonferroni<0.05) %>% 
        dplyr::select(dataSetID, shortDataSetName, inGroups, Bonferroni, FDR, enrichmentRatio, effectiveClassSize, effectiveSetSize, nCommonGenes) %>%
        arrange(Bonferroni, desc(enrichmentRatio))
  if(nrow(df)>0){
    print(kable(df, caption = paste0('Enriched terms for module ', top_modules[i], ' (SFARI Genes = ',
                                    round(SFARI_genes_by_module$perc[SFARI_genes_by_module$Module==top_modules[i]][1],4), '%)')))
    cat('\n') 
  } else cat(paste0('\nNo enrichment found for module ',top_modules[i],'\n'))
}

rm(i, df)
```
<br>

### Reactome Results

```{r, results = 'asis'}

for(i in 1:length(top_modules)){
  df = enrichment$enrichmentTable %>% filter(class==top_modules[i] & grepl('BioSystems.REACTOME',dataSetID) & Bonferroni<0.05) %>% 
        dplyr::select(dataSetID, shortDataSetName, inGroups, Bonferroni, FDR, enrichmentRatio, effectiveClassSize, effectiveSetSize, nCommonGenes) %>%
        arrange(Bonferroni, desc(enrichmentRatio))
  if(nrow(df)>0){
    print(kable(df, caption = paste0('Enriched terms for module ', top_modules[i], ' (SFARI Genes = ',
                                    round(SFARI_genes_by_module$perc[SFARI_genes_by_module$Module==top_modules[i]][1],4), '%)')))
    cat('\n') 
  } else cat(paste0('\nNo enrichment found for module ',top_modules[i],'\n'))
}

rm(i, df)
```
<br>

### GO Results

```{r, results = 'asis'}

for(i in 1:length(top_modules)){
  df = enrichment$enrichmentTable %>% filter(class==top_modules[i] & grepl('GO',dataSetID) & Bonferroni<0.05) %>% 
        dplyr::select(dataSetID, shortDataSetName, inGroups, Bonferroni, FDR, enrichmentRatio, effectiveClassSize, effectiveSetSize, nCommonGenes) %>%
        arrange(Bonferroni, desc(enrichmentRatio))
  if(nrow(df)>0){
    print(kable(df, caption = paste0('Enriched terms for module ', top_modules[i], ' (SFARI Genes = ',
                                    round(SFARI_genes_by_module$perc[SFARI_genes_by_module$Module==top_modules[i]][1],4), '%)')))
    cat('\n') 
  } else cat(paste0('\nNo enrichment found for module ',top_modules[i],'\n'))
}

rm(i, df)
```
<br>

### Internal Collection Results

```{r, results = 'asis'}

for(i in 1:length(top_modules)){
  df = enrichment$enrichmentTable %>% filter(class==top_modules[i] & !grepl('GO',dataSetID) & !grepl('BioSystems',dataSetID) & Bonferroni<0.05) %>% 
        dplyr::select(dataSetID, shortDataSetName, inGroups, Bonferroni, FDR, enrichmentRatio, effectiveClassSize, effectiveSetSize, nCommonGenes) %>%
        arrange(Bonferroni, desc(enrichmentRatio))
  if(nrow(df)>0){
    print(kable(df, caption = paste0('Enriched terms for module ', top_modules[i], ' (SFARI Genes = ',
                                    round(SFARI_genes_by_module$perc[SFARI_genes_by_module$Module==top_modules[i]][1],4), '%)')))
    cat('\n') 
  } else cat(paste0('\nNo enrichment found for module ',top_modules[i],'\n'))
}

rm(i, df)
```
<br>
